General purpose gene annotation using deep learning

### 2020-04-22 paper/main.tex
some rewrite

\subsection{Model Architecture and Usage}
\label{sec:model}
We use a 4-layer deep stacked BLSTM network with 256 units per layer and layer normalization \citep{ba2016layer} between each BLSTM layer to produce the predictions in the form of a base pair wise classification. The model consists of circa 5.4 million parameters and was implemented with the deep learning library keras \citep{chollet2015keras} on top of TensorFlow \citep{abadi2016tensorflow}. We tested multiple different architectures before arriving at this model configuration, including convolutional neural networks (CNN) and hybrid architectures. The hyperparameters were optimized by a combination of manual and automatic optimization. Automatic optimization was carried out by using either the TPE algorithm \citep{bergstra2011algorithms}, random search or grid search depending on the situation. We used NNI \citep{nni2019} to facilitate the search. All relevant hyperparameters can be found in Table \ref{tab:params} or in the Helixer source code repository.

@@

We also used multiple techniques to improve the prediction quality after the training was done. One very effective way for genomes with larger genes was to input longer sequences during inference time than during training. This was done for all animal genomes except the invertebrates. The input sequences were up to 10 times larger, depending on species group and assembly quality. This also demonstrates the ability of our model to generalize as it is able to make successful prediction on far longer sequences than it has ever seen. The sequence length during inference can be up to 200.000 base pairs long and is determined by the phylogenetic class. The concrete lengths were chosen to keep the average gene length roughly proportional to the length of the sequence input. For more implementation details on this see Section \ref{sec:longer} or the source code.

The final predictions of a single model are constructed by overlapping predictions, that were made from a sliding window and then cut to a core sequence. This was done to strongly reduce a typical drop in performance of the models towards the beginning and end of each sequence (see figure \ref{fig:length_wise_bias}) and also to improve model performance in general by providing the model with multiple different starting points. The different overlapping sequences were recombined by averaging the individual softmax values of each base. Supplementary Material (?) contains the influence on overlapping with respect to the position in a sequence for each genome, ordered by N75. We found that overlapping works best if the genomes are not very fragmented.

>>>>

\subsection{Model Architecture}
\label{sec:model}
We use a 4-layer deep stacked BLSTM network with 256 units per layer and layer normalization \citep{ba2016layer} between each BLSTM layer to produce the predictions in the form of a base pair wise classification. The model consists of circa 5.4 million parameters and was implemented with the deep learning library keras \citep{chollet2015keras} on top of TensorFlow \citep{abadi2016tensorflow}. We tested multiple different architectures before arriving at this model configuration, including convolutional neural networks (CNN) and hybrid architectures. The hyperparameters were optimized by a combination of manual and automatic optimization. Automatic optimization was carried out by using either the TPE algorithm \citep{bergstra2011algorithms}, random search or grid search depending on the situation. We used NNI \citep{nni2019} to facilitate the search. All relevant hyperparameters can be found in Table \ref{tab:params} or in the Helixer source code repository.

@@

\subsection{Inference Techniques}
We also used multiple techniques to improve the prediction quality after the training was done. One very effective way for genomes with larger genes was to input longer sequences during inference time than during training. This was done for all animal genomes except the invertebrates. The input sequences were up to 10 times larger, depending on species group and assembly quality. This also demonstrates the ability of our model to generalize as it is able to make successful prediction on far longer sequences than it has ever seen. The sequence length during inference can be up to 200.000 base pairs long and is determined by the phylogenetic class. The concrete lengths were chosen to keep the typical average gene length roughly proportional to the length of the sequence input. For more implementation details on this see Section \ref{sec:longer} or the source code.

The final predictions of a single model are constructed by overlapping predictions, that were made from a sliding window and then cut to a core sequence. This was done to strongly reduce a typical drop in performance of the models towards the beginning and end of each sequence (see Figure \ref{fig:length_wise_bias}). It also improves the average model performance in general by providing the model with multiple different starting points. The different overlapping sequences were recombined by averaging the individual softmax values of each base. The figures in Section \ref{sec:overlapping} show the effect of overlapping for each genome, ordered by N75. We found that overlapping tends to work best if the genomes are not very fragmented and we used it for both animals and plants.
